Safe Transferable Regions

### 2015-11-20 paper/type-inference.tex
Merged paper to techrep.

\name's region type system imposes a heavy annotation burden, and the
cost of manually annotating C\# standard libraries with region types
is prohibitive. We now present our region type inference algorithm
that completely eliminates the need to write region type annotations,
except on some higher-order functions. Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., \FB without region types, but with \C{letregion} and \C{open}
expressions, similar to the language introduced in
\S~\ref{sec:overview}) to programs in \FB. The key aspects of this
elaboration are described below.

% We formalize region type inference as an elaboration function from
% programs in $\absof{\FB}$ (i.e., region annotation-erased \FB) to
% programs in \FB.  The elaboration proceeds in four steps. In the first
% step we make use of the observation that region types are
% \emph{refinements} of FGJ types with region annotations and
% constraints over such region annotations, and compute polymorphic
% region type templates for methods and classes from their FGJ types.
% The templates contain region variables ($\rho$) to denote unknown
% region annotations, and predicate variables ($\varphi$) to denote
% unknown constraints over such region annotations. Free region
% variables are generalized in types (hence, polymorphic). Second, we
% elaborate expressions by introducing region variables to denote
% unknown region arguments in \C{new} expressions, method calls and
% function applications. While elaborating expressions, we make use of
% the computed region type templates to build a system of constraints
% that capture well-formedness requirements and subtyping relationships
% between type templates that must hold (as per the static semantics in
% Fig.~\ref{fb-staticsem}) for the elaboration to be valid.  Third, we
% lift expression elaboration and constraint generation to methods and
% classes. Finally, we solve the constraints to compute solutions to
% region ($\rho$) and predicate ($\varphi$) variables.
%
% Due to the presence of region-polymorphic
% recursion in \FB, constraints generated by the algorithm can be
% circular. More precisely, constraints generated can assume the form
% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
% non-idempotent substitution function for region variables in
% $\varphi$. In the constraint solving phase, the algorithm then relies
% on a fixpoint constraint solving algorithm called \csolvestar to solve
% the constraints and determine assignments for unknown region and
% predicate variables.

@@

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. FGJ type $\RgnZ\inang{T}$ is special; it is always
templatized as $\RgnZ{\inang{T}}\inang{\toprgn}$. Given the
region-annotated definition of \C{Pair} class from
\S\ref{sec:fb-syntax} a region type template for a method with FGJ
type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$ is \footnote{In
our exposition, we assume that classes $\C{A}$ and $\C{B}$ are trivial
subclasses of $\ObjZ$ with no fields/methods. Like $\ObjZ$, they
accept one region parameter - the allocation region of their
objects.}\footnote{We abuse arrow notation to also represent types of
methods, but unlike function types, there is no allocation region
annotation atop the arrow.}
$\inang{\rhoalloc_0,\rho_{1-4} \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_{1-3}} \rightarrow
\C{A}\inang{\rho_4}$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh
region variables, and $\varphi_0$ is a fresh predicate variable
denoting unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.
Region type template of a type variable $\C{a}$ is $\C{a}@\rho$, where
$\rho$ is fresh. For a class, a template is computed based on the 
type templates of its fields, constructors and methods. Additional
region and predicate variables are added as necessary. Following template 
of the \C{Pair} class illustrates this point:
% For a class, a template is computed in two steps. In
% the first step, we templatize the types of all its fields, constructor
% arguments, and arguments and return values of all its methods, along
% with the type of its superclass. In the second step, we generalize all
% the free region variables occuring in the templates computed in the
% first step as region parameters of the class. Finally, we add a fresh
% allocation region parameter ($\rhoalloc$) to the list of parameters,
% and introduce a new predicate variable ($\varphi$) to denote unknown
% constraints on region parameters. For example, consider the standard
% FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
% and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
% respectively. It can be templatized as following :
\vspace*{-0.1in}
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst; $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { 
    super(); this.fst=fst; this.snd=snd; 
  }
  a@$\rho_5$ getFst$\inang{\rhoalloc_1,\rho_5\,|\,\varphi_1}$() { return this.fst; }
}
\end{codejava}
\end{center}
\vspace*{-0.1in}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
% The template elides $\C{Pair}$'s methods, whose region type
% templates contain no free variables. Among the region parameters of
% the class template, $\rho_{0-4}$ are obtained by generalizing free
% region variables in the types of its class's fields, constructor
% arguments, and its superclass type.  The remaining parameter
% ($\rhoalloc_0$) is a fresh region variable denoting the allocation
% region argument. Fresh predicate variable $\varphi_0$ denotes
% unknown constraints over $\rhoalloc_0$ and $\rho_{0-4}$ that need to
% hold for template to be a well-formed region-annotated class
% definition in $\FB$. 
%
% The type template for a recursively defined class is computed in two
% steps. First, all recursive occurances of the class among the types of
% its fields are ignored and the class is templatized as if it is a
% non-recursive class. Next, all the recursive occurances are
% templatized with respect to the class template computed in the first
% step, such that their region annotations are \emph{exactly} same as
% the region parameters of the class. Applying this technique on a
% recursively defined $\C{LinkedListNode}\inang{\C{a}}$ class, for
% example, results in a template that uses same region variable to
% denote allocation regions of the current node, and the node that the
% current node refers to. 
%
% For example, consider a generic $\C{ListNode}\inang{\C{a}}$ class
% containing two fields: $\C{data}$ of type $\C{a}$ and $\C{next}$ of
% type $\C{ListNode}\inang{\C{a}}$.  where $\C{a}$ is the type of the
% data stored in the linked list node.  To templatize the
% $\C{ListNode}\inang{a}$ class, we first ignore its recursive
% occurance in the type of $\C{next}$ field, and templatize the type
% $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this type
% template of $\C{data}$ field, we compute the class's template as
% following: \begin{center} $\C{ListNode}\inang{\C{a}\extends\ObjZ}
% \inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
% \end{center} Next, we trivially templatize the type of $\C{next}$
% field in the body of the class as
% $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.  The
% resulting class represents a linked list with spine in the region
% $\rhoalloc_0$ and data objects in the region $\rho_0$.
% Observe that our strategy of trivially templatizing the recursive
% occurances of a class type in its definition ensures that no new
% region variables are introduced.
The type template for for a recursively defined class is computed in
two steps. First, a template is computed by ignoring the recursive
occurrences. Next, all the recursive occurrences are templatized based
on the template computed in first step.  This approach can be extended
to mutually recursive definitions in a straightforward manner. Using
the techniques outlined above, we compute region type templates for
all classes bound in the class table of the $\absof{\FB}$ program
before we proceed to elaborate expressions.

%\input{fb-elabexpr}
%\input{fb-elabmeth}

\subsection{Elaboration and Constraint Generation}

Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
(a). replacing core types occurring in expressions with fresh region
type templates, and (b). explicitly instantiating region parameters
with fresh region variables in method calls and function applications.
This elaboration is performed with respect to the polymorphic type
templates of classes and methods computed as per
\S\ref{sec:fb-templatization}. Our expression elaboration
function\footnote{Supplement contains the pseudocode for all the
elaboration functions.} traverses expressions in a syntax-directed
manner of a type checker, introducing fresh region type templates for
unknown region types, while generating constraints over region and
predicate variables. The precise nature of generated constraints is
explained in \S\ref{sec:constraints}, but in summary, they
capture the relationships between the type templates of various
subexpressions and their well-formedness.

% Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
% this elaboration for a subset of expressions in $\absof{\FB}$
% %, whose corresponding $\FB$ expressions have been ascribed static
% %semantics in Fig.~\ref{fig:fb-staticsem}. 
% $\elabExpr$ is defined under the same context as the expression typing
% judgment in Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$,
% and $\env$ retaining their meaning. The function traverses expressions
% in a syntax-directed manner of a type checker, introducing fresh
% region type templates for unknown region types, while generating
% constraints over region and predicate variables. The precise nature of
% generated constraints is explained in \S\ref{sec:fb-constraintsem},
% but in summary, they capture the relationships between the type
% templates of various subexpressions and their well-formedness. Note
% that $\elabExpr$ returns the region type template of the
% subexpression, which is used to generate constraints for the
% expression. Functions $\typeOk$ and $\subtypeOk$ (definitions not
% shown) used by $\elabExpr$ implement type well-formedness and subtype
% judgments from Fig.~\ref{fig:fb-staticsem}, respectively.

Expression elaboration can be trivially lifted to the level of a
method, and method elaboration to the level of a class, while
accumulating constraints. Besides such constraints, elaboration
functions for methods and classes also generate new constraints by
checking well-formedness and subtype relationships among the types of
their constituents. For instance, method elaboration generates a new
constraint that captures the requirement that, under the guarantees
provided by the context, the region type of a method body must be a
subtype\footnote{Subtyping is type equality in \FB. In \name,
subtyping between region types is a straightforward extension of
subtyping between core types.} of method's return type.  Likewise, the
elaboration function for a class generates a fresh constraint that
captures a couple of safety conditions: first, the allocation regions
of objects referred by the instance variables should outlive the
allocation region of the instance itself, and second, the allocation
regions of a class type and its superclass type must be the same.

Elaborating a class table ($CT$) entails elaborating every class
definition in $CT$, while collecting constraints. The constraints thus
collected are finally solved to obtain assignments for free region and
predicate variables introduced during templatization and elaboration
stages. The solutions are applied to the class table (and to the
artifacts that make up the class table, recursively) to compute a
class table that maps classes to their fully region-annotated
definitions in $\FB$.

% \subsection{Method and Class Elaboration}

% Elaborating a method involves performingexpression elaboration over
% its body, while collecting constraints. Method elaboration can be
% lifted to the level of classes

% Functions $\elabMeth$ and $\elabClass$ shown in
% Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
% class definitions, respectively. Both functions first build a context
% ($\A$) containing a set ($ \rhoenv$) of regions that are currently
% live, a map ($\aenv$) mapping type variables to their bounds, and a
% constraint formula ($\phicx$) capturing constraints over live region
% variables. We use predicate variables ($\varphi$ and $\varphi_m$) to
% capture constraints over variables in $\rhoenv$ denoting the fact that
% such constraints are yet to be inferred. Both elaboration functions
% generate new constraints by checking well-formedness and subtype
% relationships among types of their constituents. A fresh constraint
% ($C_4$) generated by $\elabClass$ captures a couple of safety
% conditions: first, the allocation regions of objects referred by the
% instance variables should outlive the allocation region of the
% instance itself, and second, the allocation regions of a class type
% and its superclass type must be the same.

% Function $\elabMeth$ elaborates a method definition of class $B$. It
% calls $\elabExpr$ with the context $\A$, its allocation context
% parameter ($\rhoallocm$), and a type environment ($\env$) that
% contains region type bindings for all the arguments of the method,
% including the implicit $\C{this}$ argument. The region type template
% returned by $\elabExpr$ for the method body is checked against its
% expected type (derived from the type template of the method)
% generating more constraints. The function then returns the elaborated
% method definition and the set of constraints.

% $\elabClass$ elaborates the definition of a class $B$. It relies on
% $\elabCons$ (not shown) and $\elabMeth$ functions to elaborate $B$'s
% constructor ($k$) and method definitions ($\bar{d}$), respectively. To
% the set of constraints returned by these functions, $\elabClass$ adds
% constraints generated by checking the well-formedness of the type
% templates of its superclass and fields, and also a new constraint
% capturing a couple of safety conditions: first, the allocation regions
% of objects referred by the instance variables should outlive the
% allocation region of the instance itself, and second, the allocation
% regions of a class type and its superclass type must be the same.

% Function $\elabClassTable$ (not shown) elaborates
% every definition in the class table $CT$, while accumulating
% constraints. The constraints are finally solved for free region and
% predicate variables introduced during templatization and elaboration
% stages. The solutions are applied to the class table (and to the
% artifacts that make up the class table, recursively) to compute a
% class table that maps classes to their fully region-annotated
% definitions in $\FB$.

% Note that if the original program in $\absof{\FB}$ contains unsafe
% references, for example, a reference from a transferable region to a
% stack regions, then the constraints generated during the elaboration
% are not satisfiable. In such case, $\solve$ fails to solve
% constraints, causing the program to be rejected.

\subsection{Constraints}
\label{sec:constraints}

Three kinds of constraints are generated during elaboration:
\begin{itemize}
\item Well-formedness constraints of form $\rho \in \rhoenv$,
restricting the domain of unification for a region variable ($\rho$)
to the set ($\rhoenv$) of regions in scope,
\item Well-formedness constraints of form
$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
variable ($\varphi$) to the set of all possible constraint formulas
over region variables ($\rhoenv$) in scope, and
\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
extended with predicate variables and \emph{pending
substitutions}\footnote{we borrowed this terminology
from~\cite{ltpldi08}}:
\begin{smathpar}
\begin{array}{lcl}
\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
    \ALT F(\varphi) \ALT \phi \conj \phi\\
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
\end{itemize}
A pending substitution ($F$) is a substitution function over region
variables/identifiers. They represent the substitutions that need to
be carried out when a predicate variable ($\varphi$) is replaced by a
concrete formula in a validity constraint. For instance, in the
validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.
% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
% In general, validity constraints generated by our algorithm assume one
% of the following two forms:
% \begin{center}
% \( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {\phicstr} \qquad
%    \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
%            {F_j(\varphi_j)}
% \)
% \end{center}
% Where $\varphi_i$'s denote the unknown preconditions of the class and
% the method under elaboration. If the constraint is generated while
% checking the well-formedness of a type or elaborating an expression,
% then $\varphi_j$'s denote the unknown preconditions of classes and
% methods that were used in that type or expression. Each use of a
% (region-polymorphic) class or a method has to instantiate region
% parameters, resulting in pending substitutions ($F_j$). Formulas
% $\phictxt$ and $\phicstr$ are concrete, i.e., free of predicate
% variables and pending substitutions. While $\phictxt$ captures
% relationships that are \emph{known} to hold between concrete region
% identifiers (i.e., $\rgn$'s) when the constraint was generated,
% $\phicstr$ captures relationships that are \emph{required} to hold
% among region varibles (i.e., $\rho$'s), or relationships between
% region variables and identifiers. Each region variable occuring in
% $\phicstr$ has an associated well-formedness constraint, which
% specifies its unification domain. The unification domain of a
% constraint is the union of unification domains of all region variables
% occuring in the constraint.

@@

A dependency DAG makes the dependencies between constraints explicit.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph's transpose.
Accordingly, we obtain a topological ordering of nodes in the graph
$G_{{C}}^{T}$ ($G_C$'s transpose), and solve the sets of constraints
in that order. The solutions obtained after solving a constraint set
are applied to the constraints in subsequent sets before attempting to
solve them. For the DAG in figure above, we consider the topological
order $[C_1, C_2, C_3]$ of its transpose, and solve the sets of
constraints in that order.

% Consequently, when the turn of a constraint set ($C$) arrives during
% the constraint solving process, it satisfies certain properties:
% \begin{itemize}
% \item There exists only one predicate variable ($\varphi$) that is either
% constrained or used by the constraints in the set ($C$). The variable is
% called set's \emph{subject}. This property follows from (a). the fact
% that all the dependency constraints have already been solved (and
% solutions applied), and (b). the assumption that there are no mutually
% recursive definitions. 
% \item All the constraints that constrain the set's subject are present
% in the set. This follows from our definition of the dependency relation.
% \end{itemize}

Solving a set ($C$) of constraints involves finding assignments for
free region ($\rho$) and predicate ($\varphi$) variables in the set.
Assignments to region variables can be found via unification (for the
examples in \S~\ref{sec:constraints}, we have already performed the
unification, so $C_{1-3}$ do not contain free region variables). Once
the region variables have been unified, we solve $C$ to find an
assignment for free predicate variables. The first step is to reduce
the set of constraints into an equivalent single constraint by
conjoining antecedents and consequents of all constraints in the
set\footnote{Supplement contains formal justification for this step}.
For the constraint sets $C_1$, $C_2$, and $C_3$, the equivalent
constraint are $c_{12}$, $c_{13}$, and $c_{14}$, respectively, shown
below:
\begin{smathpar}
\begin{array}{l}

>>>>

\name's region type system imposes a heavy annotation burden, and the
cost of manually annotating C\# standard libraries with region types
is prohibitive. We now present our region type inference algorithm
that completely eliminates the need to write region type annotations,
except on some higher-order functions. Formally, the type inference
algorithm is an elaboration function from programs in $\absof{\FB}$
(i.e., \FB without region types, but with \C{letregion} and \C{open}
expressions, similar to the language introduced in
\S~\ref{sec:overview}) to programs in \FB.  The elaboration proceeds
in four steps. In the first step we make use of the observation that
region types are \emph{refinements} of FGJ types with region
annotations and constraints over such region annotations, and compute
polymorphic region type templates for methods and classes from their
FGJ types. The templates contain region variables ($\rho$) to denote
unknown region annotations, and predicate variables ($\varphi$) to
denote unknown constraints over such region annotations. Free region
variables are generalized in types (hence, polymorphic). Second, we
make use of the computed region type templates to elaborate
expressions by introducing region variables to denote unknown region
arguments in \C{new} expressions, method calls and function
applications. While elaborating expressions, we also build a system of
constraints that capture well-formedness requirements and subtyping
relationships between type templates that must hold (as per the static
semantics in Fig.~\ref{fb-staticsem}) for the elaboration to be valid.
Third, we lift expression elaboration and constraint generation to
methods and classes. Finally, we solve the constraints by making use
of our fixpoint constraint solving algorithm \csolvestar, which
reduces the constraint solving problem to an abduction problem in a
Herbrand constraint system, and then relies on \csolve, our abduction
solver for that domain. 
% Due to the presence of region-polymorphic
% recursion in \FB, constraints generated by the algorithm can be
% circular. More precisely, constraints generated can assume the form
% $\varphi \Leftrightarrow \phi \wedge F(\varphi)$, where $F$ is a
% non-idempotent substitution function for region variables in
% $\varphi$. In the constraint solving phase, the algorithm then relies
% on a fixpoint constraint solving algorithm called \csolvestar to solve
% the constraints and determine assignments for unknown region and
% predicate variables.

@@

Region type templates are FGJ types extended with fresh region
variables ($\rho$) and predicate variables ($\varphi$) to denote
unknown region annotations and region constraints, respectively. For
instance, if a variable $x$ has type $\ObjZ$ in FGJ, its region type
template is of form $\ObjZ\inang{\rho_0}$, where $\rho_0$ is a fresh
region variable. Likewise, given the region-annotated definition of
\C{Pair} class from \S\ref{sec:fb-syntax} a region type template for a
method with FGJ type $\C{Pair}\inang{\C{A},\C{B}} \rightarrow \C{A}$
is \footnote{In our exposition, we assume that classes $\C{A}$ and
$\C{B}$ are trivial subclasses of $\ObjZ$ with no fields/methods. Like
$\ObjZ$, they accept one region parameter - the allocation region of
their objects.}\footnote{We abuse arrow notation to also represent
types of methods, but unlike function types, there is no allocation
region annotation atop the arrow in a method type.}
$\inang{\rhoalloc_0,\rho_1,\rho_2,\rho_3 \,|\, \varphi_0}
\C{Pair}\inang{\C{A},\C{B}} \inang{\rho_1,\rho_2,\rho_3} \rightarrow
\unitZ$, where $\rhoalloc_0$ and $\rho_{1-3}$ are fresh region
variables, and $\varphi_0$ is a fresh predicate variable denoting
unknown constraints over $\rhoalloc_0$ and $\rho_{1-3}$.  Region type
template of a type variable $\C{a}$ is $\C{a}@\rho$, where $\rho$ is
fresh. For a class, a template is computed in two steps. In the first
step, we templatize the types of all its fields, constructor
arguments, and arguments and return values of all its methods, along
with the type of its superclass. In the second step, we generalize all
the free region variables occuring in the templates computed in the
first step as region parameters of the class. Finally, we add a fresh
allocation region parameter ($\rhoalloc$) to the list of parameters,
and introduce a new predicate variable ($\varphi$) to denote unknown
constraints on region parameters. For example, consider the standard
FGJ definition of $\C{Pair}\inang{\C{a},\C{b}}$ class, where $\C{a}$
and $\C{b}$ are the types of $\C{fst}$ and $\C{snd}$ fields,
respectively. It can be templatized as following:
\begin{center}
\begin{codejava}
class Pair$\inang{a\extends\ObjZ, b\extends\ObjZ}$
          $\inang{\rhoalloc_0,\rho_{0-4}\,|\,\varphi_0} \extends \ObjZ\inang{\rho_4}$ {
  $a@\rho_0$ fst;
  $b@\rho_1$ snd;
  Pair($a@\rho_2$ fst, $b@\rho_3$ snd) { $...$ }
  $...$
}
\end{codejava}
\end{center}
% \footnote{In reality, the generalized region variables only come
% from the type templates of class's fields and superclass. Method
% types do not contain any free region variables.}
The template elides $\C{Pair}$'s methods, whose region type templates
contain no free variables. Among the region parameters of the class
template, $\rho_{0-4}$ are obtained by generalizing free region
variables in the types of its class's fields, constructor arguments,
and its superclass type.  The remaining parameter ($\rhoalloc_0$) is a
fresh region variable denoting the allocation region argument. Fresh
predicate variable $\varphi_0$ denotes unknown constraints over
$\rhoalloc_0$ and $\rho_{0-4}$ that need to hold for template to be a
well-formed region-annotated class definition in $\FB$.

The type template for a recursively defined class is computed in two
steps. First, all recursive occurances of the class among the types of
its fields are ignored and the class is templatized as if it is a
non-recursive class. Next, all the recursive occurances are
templatized with respect to the class template computed in the first
step, such that their region annotations are exactly same as the
region parameters of the class. For example, consider a generic
$\C{ListNode}\inang{\C{a}}$ class containing two fields: $\C{data}$ of
type $\C{a}$ and $\C{next}$ of type $\C{ListNode}\inang{\C{a}}$.
where $\C{a}$ is the type of the data stored in the linked list node.
To templatize the $\C{ListNode}\inang{a}$ class, we first ignore its
recursive occurance in the type of $\C{next}$ field, and templatize
the type $\C{a}$ of $\C{data}$ field as $\C{a}@\rho_0$. Based on this
type template of $\C{data}$ field, we compute the class's template as
following:
\begin{center}
$\C{ListNode}\inang{\C{a}\extends\ObjZ}
\inang{\rhoalloc_0,\rho_0\,|\,\varphi_0} \extends \ObjZ \{...\}$
\end{center}
Next, we trivially templatize the type of $\C{next}$ field in the body
of the class as $\C{ListNode}\inang{\C{a}}\inang{\rhoalloc_0,\rho_0}$.
The resulting class represents a linked list with spine in the region
$\rhoalloc_0$ and data objects in the region $\rho_0$.

% Observe that our strategy of trivially templatizing the recursive
% occurances of a class type in its definition ensures that no new
% region variables are introduced.
The templatization technique we described above for recursive class
definitions can be extended to mutually recursive definitions in a
straightforward manner, by simultaneously templatizing them. Using the
techniques outlined above, we compute region type templates for all
classes bound in the class table of the $\absof{\FB}$ program before
we proceed to elaborate expressions.

\input{fb-elabexpr}
\input{fb-elabmeth}

\subsection{Expression Elaboration}

Elaborating $\absof{\FB}$ expressions to $\FB$ expressions involves
(a). replacing core types in variable declarations and \C{new}
expressions with fresh region type templates, and (b). explicitly
instantiating region parameters of methods with fresh region variables
in method calls and function applications. This elaboration is
performed with respect to the polymorphic type templates of classes
and methods computed as per \S\ref{sec:fb-templatization}.

Function $\elabExpr$, shown in Fig.~\ref{fig:fb-elabexpr}, performs
this elaboration for a subset of expressions in $\absof{\FB}$, whose
corresponding $\FB$ expressions have been ascribed static semantics in
Fig.~\ref{fig:fb-staticsem}. $\elabExpr$ is defined under the same
context as the expression typing judgment in
Fig.~\ref{fig:fb-staticsem} with symbols $\A$,$\ralloc$, and $\env$
retaining their meaning. The function traverses expressions in a
syntax-directed manner of a type checker, introducing fresh region
type templates for unknown region types, while generating constraints
over region and predicate variables. The precise nature of generated
constraints is explained in \S\ref{sec:fb-constraintsem}, but in
summary, they capture the relationships between the type templates of
various subexpressions and their well-formedness. Note that
$\elabExpr$ returns the region type template of the subexpression,
which is used to generate constraints for the expression. Functions
$\typeOk$ and $\subtypeOk$ (definitions not shown) used by $\elabExpr$
implement type well-formedness and subtype judgments from
Fig.~\ref{fig:fb-staticsem}, respectively.

\subsection{Method and Class Elaboration}

Functions $\elabMeth$ and $\elabClass$ shown in
Fig.~\ref{fig:fb-elabmeth} lift expression elaboration to method and
class definitions, respectively. Both functions first build a context
($\A$) containing a set ($ \rhoenv$) of region variables denoting
regions that are currently live, a map ($\aenv$) mapping type
variables to their bounds, and a constraint formula ($\phicx$)
capturing constraints over live region variables. We use predicate
variables ($\varphi$ and $\varphi_m$) to capture constraints over
variables in $\rhoenv$ denoting the fact that such constraints are yet
to be inferred.

Function $\elabMeth$ elaborates a method definition of class $B$. It
calls $\elabExpr$ with the context $\A$, its allocation context
parameter ($\rhoallocm$), and a type environment ($\env$) that
contains region type bindings for all the arguments of the method,
including the implicit $\C{this}$ argument. The region type template
returned by $\elabExpr$ for the method body is checked against its
expected type (derived from the type template of the method)
generating more constraints. The function then returns the elaborated
method definition and the set of constraints.

$\elabClass$ elaborates the definition of a class $B$. It relies on
$\elabCons$\footnote{The definition of $\elabCons$ is straightforward,
hence not shown.} and $\elabMeth$ functions to elaborate $B$'s
constructor ($k$) and method definitions ($\bar{d}$), respectively. To
the set of constraints returned by these functions, $\elabClass$ adds
constraints generated by checking the well-formedness of the type
templates of its superclass and fields, and also a new constraint
capturing a couple of safety conditions: first, the allocation regions
of objects referred by the instance variables should outlive the
allocation region of the instance itself, and second, the allocation
regions of a class type and its superclass type must be the same.

Function $\elabClassTable$ (Fig.~\ref{fig:fb-elabmeth}) elaborates
every definition in the class table $CT$, while accumulating
constraints. The constraints are finally solved by $\solve$
(\S\ref{sec:fb-constraintsolving}), which returns substitution
functions $\substFn_\rho$ and $\substFn_\varphi$ for free region and
predicate variables, respectively, introduced during templatization
and elaboration stages. The substitutions are applied to the class
table (and to the artifacts that make up the class table, recursively)
to compute a class table that maps classes to their fully
region-annotated definitions in $\FB$.

Note that if the original program in $\absof{\FB}$ contains unsafe
references, for example, a reference from a transferable region to a
stack regions, then the constraints generated during the elaboration
are not satisfiable. In such case, $\solve$ fails to solve
constraints, causing the program to be rejected.

\subsection{Constraints}
\label{sec:fb-constraintsem}

Our constraint generation algorithm generates three kinds of
constraints:
\begin{itemize}
\item Well-formedness constraints of form $\rho \in \rhoenv$,
restricting the domain of unification for a region variable ($\rho$)
to the set ($\rhoenv$) of regions in scope,
\item Well-formedness constraints of form
$\tywf{\rhoenv}{\varphi}$, restricting the domain of a predicate
variable ($\varphi$) to the set of all possible constraint formulas
over region variables ($\rhoenv$) in scope, and
\item Validity constraints of form $\isvalid{\phicx}{\phi}$, where
$\phicx$ and $\phi$ are region constraints (Fig.~\ref{fig:fb-syntax})
extended with predicate variables and \emph{pending
substitutions}\footnote{we borrowed this terminology
from~\cite{ltpldi08}}:
\begin{smathpar}
\begin{array}{lcl}
\phicx,\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho 
    \ALT F(\varphi) \ALT \phi \conj \phi\\
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
\end{itemize}
A pending substitution ($F$) is a substitution function over region
variables/identifiers. They represent the substitutions that need to
be carried out when a predicate variable ($\varphi$) is replaced by a
concrete formula in a validity constraint. For instance, in the
validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.

% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises. 
In general, validity constraints generated by our algorithm assume one
of the following two forms:
\begin{center}
\( \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {\phicstr} \qquad
   \isvalid{\phictxt \conj \bigwedge_{i}\varphi_i}
           {F_j(\varphi_j)}
\)
\end{center}
Where $\varphi_i$'s denote the unknown preconditions of the class and
the method under elaboration. If the constraint is generated while
checking the well-formedness of a type or elaborating an expression,
then $\varphi_j$'s denote the unknown preconditions of classes and
methods that were used in that type or expression.  Each use of a
(region-polymorphic) class or a method may instantiate region
parameters differently, resulting in a different pending substitution
($F_j$). Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
of predicate variables and pending substitutions. While $\phictxt$
captures relationships that are \emph{known} to hold between concrete
region identifiers (i.e., $\rgn$'s) when the constraint was generated,
$\phicstr$ captures relationships that are \emph{required} to hold
among region varibles (i.e., $\rho$'s), or relationships between
region variables and identifiers. Each region variable occuring in
$\phicstr$ has an associated well-formedness constraint, which
specifies its unification domain. The unification domain of a
constraint is the union of unification domains of all region variables
occuring in the constraint.

@@

A dependency DAG makes the dependencies between constraints explicit.
Constraints in each set are mutually dependent, and need to be solved
simultaneously, whereas constraints in different sets can be solved as
per any valid topological ordering of the graph's transpose.
Accordingly, we obtain a topological ordering of nodes in the graph
$G_{{C}}^{T}$ ($G_C$'s transpose), and solve the sets of constraints
in that order. The solutions obtained after solving a constraint set
are applied to the constraints in subsequent sets before attempting to
solve them.  Consequently, when the turn of a constraint set ($C$)
arrives during the constraint solving process, it satisfies certain
properties:
\begin{itemize}
\item There exists only one predicate variable ($\varphi$) that is either
constrained or used by the constraints in the set ($C$). The variable is
called set's \emph{subject}. This property follows from (a). the fact
that all the dependency constraints have already been solved (and
solutions applied), and (b). the assumption that there are no mutually
recursive definitions. 
\item All the constraints that constrain the set's subject are present
in the set. This follows from our definition of the dependency relation.
\end{itemize}

For the DAG in figure above, we consider the topological order $[C_1,
C_2, C_3]$ of its transpose, and solve the sets of constraints in that
order.  Solving the set ($C$) of constraints entails finding an
assignment for $C$'s subject ($\varphi$), and also all the region
variables ($\rhobar$) that occur free in $C$, such that the solution
satisfies well-formedness constraints on $\varphi$ and $\rhobar$. To
simplify presentation, we think of $C$ as being parameterized on
$\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
\rbrack$. We now formalize the constraint satisfaction problem, and
its solution.
